Issues in developing mulitvariable models for treatment selection

### 2016-02-19 Documents/paper.Rmd
lisa review
pca selection is a bust

---
title: "Issues in developing mulitvariable models for treatment selection"
author: "Michael C Sachs and Lisa M McShane"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
    pdf_document:
      pandoc_args: ["--bibliography", "bookchapter-1.bib"]
---

@@

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE)
```

@@

A biomarker signature can inform clinical practice in a number of ways. 
In what ways can it change clinical practice?

@@

The goal of the @zhu2010prognostic paper was to identify a multi-gene signature that strongly predicts prognosis, and the hypothesis was that the poor prognosis subgroup would benefit more from ACT compared to the good prognosis subgroup. The annotated gene expression data and clinical information are available from the Gene Expression Omnibus (identifier: GSE14814, @edgar2002gene). Batch effects were removed using the ComBat function in the sva R package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations. We reproduce their analysis while illustrating the key issues under discussion.

@@

## Estimating Prediction Error

```{r funcs}
print("dick butt")
```

Recall that the main goal is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected error on future observations for $f \in \mathcal{F}$. This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$. However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large. This is because $\phi$ depends on $f$, and thus the statistic $\phi$ is being adaptively defined based on the observed data $S$, hence causing the overfitting.

@@

A variation on the cross-validation approach is bootstrapping. In that case, a sample $S_b$ of size $n$ is sampled _with replacement_ from $S$. Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$. The error $\phi$ is estimated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$. This process is repeated $K$ times to yield $K$ estimates. @efron1997improvements showed that these estimates are still biased due to the sampling with replacement. Thus they propose the 0.632 estimate: 
$$
\hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
$$
where $\hat{E}[\phi_{f}(S)]$ is the naive estimate of $\phi_f$ using the entire dataset.

@@

We compare each of the three methods described above along with the biased approaches of using the full sample to select the features, followed by either fitting the model on a split sample, or using fitting the full model inside the cross validation step. Note that @zhu2010prognostic used the latter approach, which explains the differences in our results. Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.

## Evaluating Clinical Benefit

While a biomarker signature may accurately predict a clinical outcome, that does not neccessarily imply that the signature is clinically useful.

### Developing a biomarker signature in a single-arm trial

Classify the binary response to treatment.

>>>>

---
title: "Issues in developing mulitvariable models for treatment selection"
author: "Michael C Sachs and Lisa M McShane"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
    pdf_document:
      pandoc_args: ["--bibliography", "bookchapter-1.bib"]
fontsize: 11pt
geometry: margin=1in
---

\linespread{2}

@@

```{r setup, include = FALSE}
library(ggplot2)
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE)
knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})
```

@@

A biomarker signature can inform clinical practice in a number of ways, regardless of how the signature was developed. A highly prognostic signature may identify a subpopulation that has such a good chance of long term survival, that they do not need to undergo treatment that carries risks and side effects such as chemotherapy. In the context of a specific therapy that targets a particular molecular pathway, a signature may identify a subpopulation that does not benefit from that therapy, thereby guiding the decision to treat or not. A signature that was developed to identify intrinsic molecular subtypes could be both prognostic and informative about the effectiveness of a therapy.

The key point is that signatures are useful if they can correctly and reliably classify patients into distinct subgroups for which different treatment decisions would be made. There are two distinct but related statistical concepts involved here: calibration or accuracy, and discrimination. Signatures are often optimized to be well-calibrated, that is, highly accurate for predicting outcomes. However if the signature does not separate a population into distinct subgroups, then it is unlikely to be informative enough to change clinical practice.

In the development process it is important to evaluate both of these statistical concepts. Furthermore, it is not trivial to assess each of these in a valid manner when the data are used to define the signature itself. We illustrate the potential for bias, and remedies, in our examples.

@@

The goal of the @zhu2010prognostic paper was to identify a multi-gene signature that strongly predicts prognosis, and the hypothesis was that the poor prognosis subgroup would benefit more from ACT compared to the good prognosis subgroup. The signature was trained to predict disease specific survival. The annotated gene expression data and clinical information are available from the Gene Expression Omnibus (identifier: GSE14814, @edgar2002gene). Batch effects were removed using the ComBat function in the sva R package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations. We reproduce their analysis while illustrating the key issues under discussion.

The authors of @zhu2010prognostic present results that mainly focus on the discrimination ability of their estimated signature. They do that by demonstrating that the two risk subgroups predicted by their signature (high risk and low risk) have separation in their survival curves and that the hazard ratio for their signature is large and significant even when adjusting for other risk factors. They do not directly address calibration, that is, whether their signature accurately predicts survival times.

@@

Recall that the main goal is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected value of a given statistic on future observations for $f \in \mathcal{F}$. This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$. However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large. This is because $\phi$ depends on $f$, and thus the statistic $\phi$ is being adaptively defined based on the observed data $S$, hence causing the overfitting.

In many cases during signature development, the statistic $\phi$ in question is a statistic that relates to calibration, such as classification accuracy, correlation or mean squared error. While a biomarker signature may accurately predict a clinical outcome, that does not neccessarily imply that the signature is clinically useful. As we mentioned before, two statistical critera are necessary for determining clinical usefulness: calibration and discrimination. To assess discrimination, a different statistic $\phi$ may be used, such as an odds ratio, hazard ratio, or difference in survival probabilities. Evaluation of $\phi$ is also subject to bias due to overfitting, a fact that is commonly overlooked in the medical literature.

In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups as defined by the JBL.10 signature $f$. The risk groups were determined by the signature $f$, which was estimated using $S$ the same data that is then used to estimate the hazard ratio which is estimated to be 15. They go on to assess the discrimination of the JBL.10 signature in a series of independent data sets, in which they find hazard ratios around 2. We will reanalyze the dataset and illustrate some remedies for avoiding bias in determining the discrimination (an additionally the calibration) of this signature.

@altman2000we, @buyse2007towards, @moons2012riskI, @moons2012riskII

## Remedies to Overfitting

```{r funcs}
print("Insert simulation results")
```

@@

A variation on the cross-validation approach is bootstrapping. In that case, a sample $S_b$ of size $n$ is sampled _with replacement_ from $S$. Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$. The error $\phi$ is estimated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$. This process is repeated $K$ times to yield $K$ estimates. These $K$ estimates are averaged to obtain the mean over the bootstrap replicates. @efron1997improvements suggest a variation, the 0.632 estimate: 
$$
\hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
$$
where $\hat{E}[\phi_{f}(S)]$ is the naive estimate of $\phi_f$ using the entire dataset.

@@

We compare each of the three methods described above along with the biased approaches of using the full sample to select the features, followed by either fitting the model on a split sample, or using fitting the full model inside the cross validation step. Note that @zhu2010prognostic used the latter approach, which explains the differences in our results. Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$. In our simulation, we look at two different statistics, classification accuracy to assess calibration and the odds ratio for the outcome comparing the signature groups to assess discrimination.

## Evaluating Calibration

## Evaluating Discrimination and Clinical Benefit
